---
title: "Using ritalic to Access Lichen Data from ITALIC"
author: "Matteo Conti"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ritalic to Access Lichen Data from ITALIC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Introduction

The `ritalic` package provides an R interface to [ITALIC](https://italic.units.it/), the Information System on Italian Lichens. ITALIC is a comprehensive database containing information about lichen species found in Italy and neighboring European countries. It's maintained by the research group of Professor Pier Luigi Nimis at the University of Trieste.

`ritalic` allows you to access a wealth of lichen data directly from R, including:

-   **Taxonomic information:** Classification, accepted names, synonyms, etc.
-   **Geographic distribution:** Presence/absence in Italian regions and ecoregions.
-   **Ecological data:** Substrate preferences, photobiont types, growth forms, etc.
-   **Morphological traits:** Descriptions, indicator values, reproductive strategies.
-   **Occurrence records:** Herbarium specimen data, including coordinates, dates, and collectors.
-   **Identification keys:** Generate interactive keys for specific sets of species.

This vignette will guide you through the main functions of `ritalic`, demonstrating how to use them to retrieve and analyze lichen data for your research.

# Installation

```{r, eval = FALSE}
# Install the development version from GitHub:
install.packages("devtools")
devtools::install_github("plant-data/ritalic")
```

```{r}
# Load the package
library(ritalic)
```

# Matching Scientific Names: `italic_match()`

Before retrieving data for specific lichens, it's essential to ensure you're using the correct, accepted names according to ITALIC. The `italic_match()` function helps with this. It takes a character vector of scientific names and matches them against the ITALIC database, handling synonyms and infraspecific taxa.

```{r}
# Example: Matching a few names
my_names <- c("Cetraria islandica", "Xantoria parietina", "Cladonia rangiferina", "Pseudevernia furfuracea b) ceratea")
matched_names <- italic_match(my_names, subsp_marks = "b)")
print(matched_names)
```

**Explanation:**

-   `sp_names`: The vector of names you want to match.
-   `subsp_marks`, `var_marks`, `form_marks`: Optional arguments to specify uncommon infraspecific rank markers (if not using the standard "subsp.", "var.", "f.").
-   The output is a data frame with the original input name, the matched name, the nomenclatural status ("accepted" or "synonym"), the currently accepted name (if it's a synonym), and matching scores for the name and authority parts.

**Best Practice:** Always use `italic_match()` first to validate and obtain accepted names before using them in other `ritalic` functions.

# Retrieving Data

The following sections describe functions for retrieving different types of data from ITALIC. We'll use the accepted names obtained from `italic_match()` in the examples.

## 1. National Checklist: `italic_checklist()`

This function retrieves the complete list of accepted scientific names in the ITALIC checklist.

```{r}
# Get the entire checklist
all_lichens <- italic_checklist()

# Display the first few names
head(all_lichens)
```

**Note:** The checklist is large, so it's often more practical to work with subsets of species.

## 2. Taxonomic Classification: `italic_taxonomy()`

Get the taxonomic hierarchy (phylum, class, order, family, genus) for a set of accepted names.

```{r}
# Get taxonomy for our matched names
taxonomy <- italic_taxonomy(matched_names$accepted_name)
print(taxonomy)
```

## 3. Descriptions: `italic_description()`

Retrieve morphological descriptions and taxonomic/ecological notes.

```{r}
# Get descriptions
descriptions <- italic_description(matched_names$accepted_name)
print(descriptions)
```

## 4. Ecology and Traits: `italic_ecology_traits()`

Access a wide range of ecological and trait data, including substrate, photobiont, growth form, indicator values, and more.

```{r}
# Get ecology and traits
ecology_traits <- italic_ecology_traits(matched_names$accepted_name)
print(ecology_traits)
```

## 5. Distribution in Italian Regions: `italic_regions_distribution()`

Get presence/absence data for each species in the administrative regions of Italy.

```{r}
# Get regional distribution
regions_dist <- italic_regions_distribution(matched_names$accepted_name)
print(regions_dist)
```

## 6. Distribution in Ecoregions: `italic_ecoregions_distribution()`

Get distribution data across Italian ecoregions, either as commonness/rarity categories or as presence/absence (0/1).

```{r}
# Get ecoregion distribution (commonness/rarity)
ecoregions_dist_rarity <- italic_ecoregions_distribution(matched_names$accepted_name)
print(ecoregions_dist_rarity)

# Get ecoregion distribution (presence/absence)
ecoregions_dist_pa <- italic_ecoregions_distribution(matched_names$accepted_name, result_data = "presence-absence")
print(ecoregions_dist_pa)
```

## 7. Occurrence Records: `italic_occurrences()`

Retrieve herbarium specimen data, including coordinates, collection date, substrate, and institution code. You can get either "simple" or "extended" output (the latter includes locality and catalog number).

```{r}
# Get occurrence records (simple)
occurrences_simple <- italic_occurrences(matched_names$accepted_name)
head(occurrences_simple) # Display the first few records

# Get occurrence records (extended)
occurrences_extended <- italic_occurrences(matched_names$accepted_name, result_data = "extended")
head(occurrences_extended)
```

## 8. Occurrence References: `italic_occurrences_references()`

Get bibliographic references and DOIs for the publications associated with occurrence datasets. You need to first retrieve occurrence data using `italic_occurrences()`.

```{r}
# Get references for the occurrence records
references <- italic_occurrences_references(occurrences_simple)
print(references)
```

## 9. Interactive Identification Keys: `italic_identification_key()`

Generate a unique URL to an interactive dichotomous key for a set of species. This uses the KeyMaker system on the ITALIC website.

```{r, eval = FALSE}
# Generate an identification key URL
key_url <- italic_identification_key(matched_names$accepted_name)
print(key_url) # This will print the URL. You can open it in a browser.
```

Note: the link generated by this function has an expiration time.

## 10. Get all taxon data: `italic_taxon_data()`

This function returns taxonomy, ecology, and distribution data of the lichen species passed as input.

```{r}
taxon_data <- italic_taxon_data(matched_names$accepted_name)
print(taxon_data)
```

## 11. Get a presence-absence matrix of traits: `italic_traits_pa()`

This function returns a presence-absence matrix of the functional traits of the lichen species passed as input.

```{r}
traits_pa <- italic_traits_pa(matched_names$accepted_name)
print(traits_pa)
```

# Advanced Usage

## Handling API Errors and Rate Limits

The ITALIC API may return errors if there are issues with the request or if you exceed the rate limit (requests per unit of time). `ritalic` functions handle these gracefully:

-   **Error Handling:** If the API returns an error (e.g., status code 500), the functions will stop and display an informative message.
-   **Rate Limiting:** If you hit the rate limit (status code 429), the functions will automatically pause for 10 seconds before retrying (up to 5 attempts). You'll see a message indicating the pause.

## Progress Bars

For functions that may take some time to retrieve data (especially when dealing with many species), `ritalic` displays progress bars in the console so you can monitor the progress. These progress bars are only shown in interactive R sessions.

# Conclusion

`ritalic` provides a powerful and user-friendly way to access the rich lichen biodiversity data available in ITALIC. By combining the functions described in this vignette, you can perform a wide range of analyses, from basic taxonomic queries to complex ecological and biogeographic studies. Remember to consult the function documentation (`?function_name`) for more details on specific parameters and options.

If you encounter any bugs or have suggestions for improvements, please feel free to open an issue on the [GitHub repository](https://github.com/plant-data/ritalic). Happy lichen exploring!
